Intersection Types, Sub-formula Property, and the Functional Character of the Lambda Calculus

### 2018-08-09 DenotSem/determ.tex
longer proof for nat case

Armed with a good definition of intersection types, I turned back to
the proof of the Join Theorem, but first I needed to adjust my
denotational semantics to use intersection types. For this we'll need
the definition of well formed types that we alluded to earlier.

\noindent \textbf{Proposition} (Consistency is Symmetric and Reflexive) \\
%
\begin{enumerate}
\item If $A \sim B$, then $B \sim A$.
\item If $\WF{A}$, then $A \sim A$.
\end{enumerate}

@@

\noindent \textbf{Lemma} (Subtyping Inversion for Funcion Types)\\
%
If $C <: A \to B$, then there is a sequence of function types $\Gamma$
such that
\begin{enumerate}
\item each element of $\Gamma$ is an atom of $C$,
\item For every $D\to E \in \Gamma$, we have $A <: D$, and
\item $\bigsqcap \mathrm{cod}(\Gamma) <: B$.
\end{enumerate}
Note that item 2 above implies that $A <: \bigsqcap \mathrm{dom}(\Gamma)$.

With the above lemmas, we are ready to prove the Consistency and
Subtyping Lemma that previously stumped me.\\

@@

(1) The proof is by strong induction on the sum of the depths of $A$,
$B$, $C$, and $D$. To show that $C \sim D$ it suffices to show that
all of their atoms are consistent. Suppose $C' \in \ATOMS{C}$
and $D'\in\ATOMS{D}$. So we need to show that $C' \sim D'$.
%Note that $A <: C'$ and $B <: D'$
We proceed by cases on $C'$.
\begin{itemize}
\item Case $C'=n_1$:\\
  We have $A <: C'$ and therefore $n_1 \in \ATOMS{A}$.
  Then because $A \sim B$, we have $\ATOMS{B} \subseteq \{n_1\}$.
  We have $B <: D'$, so we also have $\ATOMS{D} \subseteq \{n_1\}$.
  Therefore $C' \sim D'$.

@@

The next two lemmas follow from the Consistency and Subtyping Lemma
and help prepare to prove the case for application in the Join
Theorem.

@@

We move on to the Weakening Lemma, which will be needed to prove the
Join Theorem. \\

@@

Now for the main event, the proof of the Join Theorem!
Though we'll rebrand it as the Meet Theorem.\\

\noindent \textbf{Theorem} (Meet)\\
If $A_1 \in \ESEM{e}\Gamma_1$, $A_2 \in \ESEM{e}\Gamma_2$,
both $\Gamma_1$ and $\Gamma_2$ are well formed,
and $\Gamma_1 \sim \Gamma_2$, \\
then $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1\sqcap\Gamma_2)$. \\
%
\textbf{Proof}
We proceed by induction on $e$.
\begin{itemize}
\item Case $e=x$:
  We have $\Gamma_1[k] <: A_1$ and $\Gamma_2[k] <: A_2$.

  nts. $\Gamma_1[k] \sqcap \Gamma_2[k] <: A_1 \sqcap A_2$
\end{itemize}

>>>>

We define subtyping as follows.
\[
  A <: B \quad = \quad A \vdash B
\]

The BCD subtyping rules and other derived rules follow from the above
lemmas.\\

\noindent \textbf{Proposition} (Properties of Subtyping) 
\begin{enumerate}
\item $A <: A$.
\item If $A <: B$ and $B <: C$, then $A <: C$. 
\item If $C <: A$ and $B <: D$, then $A \to B <: C \to D$.
\item If $A_1 <: B$, then $A_1 \sqcap A_2 <: B$.
\item If $A_2 <: B$, then $A_1 \sqcap A_2 <: B$.
\item If $B <: A_1$ and $B <: A_2$, then $B <: A_1 \sqcap A_2$.
\item If $A <: C$ and $B <: D$, then $A \sqcap B <: C \sqcap D$.
\item $(A\to B) \sqcap (A \to C) <: A \to (B \sqcap C)$.
\item $(A \to C) \sqcap (B \to D) <: (A\sqcap B) \to (C \sqcap D)$
\end{enumerate}

\section{Consistency and Subtyping, Resolved}

Recall that my switch to intersection types was motivated by my
failure to prove the Consistency and Subtyping Lemma.  We now return
to the proof of that Lemma. We start with a handful of lemmas that are
needed for that proof. \\

\noindent \textbf{Lemma} (Consistency is Symmetric and Reflexive) 
%
\begin{enumerate}
\item If $A \sim B$, then $B \sim A$.
\item If $\WF{A}$, then $A \sim A$.
\end{enumerate}

@@

\noindent \textbf{Lemma} (Subtyping Inversion for Function Types)\\
%
If $C <: A \to B$, then there is a sequence of function types $\Gamma$
such that
\begin{enumerate}
\item each element of $\Gamma$ is an atom of $C$,
\item For every $D\to E \in \Gamma$, we have $A <: D$, and
\item $\bigsqcap \mathrm{cod}(\Gamma) <: B$.
\end{enumerate}
Note that item 2 above implies that $A <: \bigsqcap \mathrm{dom}(\Gamma)$.

@@

(1) The proof is by strong induction on the sum of the depths of $A$,
$B$, $C$, and $D$. We define the depth of a type as follows.
\begin{align*}
  \mathit{depth}(n) &= 0 \\
  \mathit{depth}(A \to B) &= 1 + \mathrm{max}(\mathit{depth}(A),\mathit{depth}(B)) \\
  \mathit{depth}(A \sqcap B) &= \mathrm{max}(\mathit{depth}(A),\mathit{depth}(B)) 
\end{align*}
To show that $C \sim D$ it suffices to show that
all of their atoms are consistent. Suppose $C' \in \ATOMS{C}$
and $D'\in\ATOMS{D}$. So we need to show that $C' \sim D'$.
%Note that $A <: C'$ and $B <: D'$
We proceed by cases on $C'$.
\begin{itemize}
\item Case $C'=n_1$:\\
  We have $A <: C'$ and therefore $n_1 \in \ATOMS{A}$.
  Then because $A \sim B$, we have $\ATOMS{B} \subseteq \{n_1\}$.
  We have $B <: D'$, so we also have $\ATOMS{D} \subseteq \{n_1\}$.
  Therefore $C' \sim D'$.

@@

The next two lemmas follow from the Consistency and Subtyping Lemma
and help prepare to prove the case for application in the Join
Theorem. \\

@@

\section{Updating the Denotational Semantics}

Armed with the Consistency and Subtyping Lemma, I turned back to the
proof of the Join Theorem, but first I needed to update my
denotational semantics to use intersection types instead of
values. For this we'll need the definition of well formed types that
we alluded to earlier.

It is easy to show that swapping in a ``super'' environment does not
change the semantics.

@@

Now for the main event, the proof of the Join Theorem!  Though we'll
rebrand it as the Meet Theorem, given our switch from values to
intersection types.\\

\noindent \textbf{Theorem} (Meet)\\
If $A_1 \in \ESEM{e}\Gamma_1$, $A_2 \in \ESEM{e}\Gamma_2$,
both $\Gamma_1$ and $\Gamma_2$ are well formed,
and $\Gamma_1 \sim \Gamma_2$, \\
then $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1\sqcap\Gamma_2)$. \\[1ex]
%
\textbf{Proof}
We proceed by induction on $e$.
\begin{itemize}
\item Case $e=k$ ($k$ is a de Bruijn index for a variable):\\
  We have $\Gamma_1[k] <: A_1$ and $\Gamma_2[k] <: A_2$,
  so $\Gamma_1[k] \sqcap \Gamma_2[k] <: A_1 \sqcap A_2$.
  Also, because $\Gamma_1 \sim \Gamma_2$ we have
  $\Gamma_1[k] \sim \Gamma_2[k]$ and therefore
  $A_1 \sim A_2$, by the Consistency and Subtyping Lemma.
  So we have $\WF{A_1 \sqcap A_2}$ and this case is finished.

\item Case $e=n$: \\
  We have $n <: A_1$ and $n <: A_2$, so $n <: A_1 \sqcap A_2$.
  Also, we have $A_1 \sim A_2$ by the Consistency and Subtyping Lemma.
  So we have $\WF{A_1 \sqcap A_2}$ and this case is finished.

\item Case $e=\lambda e$: \\

\end{itemize}
